---
title: "NC Lead Testing Study (Analysis)"
author: "Linnea Wilson and Nate Jones"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true # table of content true
    toc_depth: 2  # upto three depths of headings (specified by #, ## and ###)
    toc_float: true
---

The goal of this page is to present analyses conducted for our upcoming publication on lead testing in North Carolina. Please email Linnea or Nate if you have any questions!

# Step 1: Setup workspace

```{r, setup, warning=FALSE,message=FALSE,error=FALSE}

#Clear Memory
remove(list = ls())

#load libraries of interest
library(tidyverse)
library(readxl)
library(ggplot2)

#read data
data_dir <- "C:/WorkspaceR/NC_Lead_Sampling/data/"
df_names <- read_xlsx(
              path = paste0(data_dir,"nc_database.xlsx"), 
              sheet =2 , 
              skip =1, 
              n_max=1) %>% 
            colnames()
df <- read_xlsx(
        path = paste0(data_dir,"nc_database.xlsx"), 
        sheet = 2, 
        skip=3, 
        col_types = c('text',rep('numeric',17)),
        col_names = df_names)
```

# Step 2: MRL handling

First step -- we need to deal with values below the minimum reporting limit. The way the data is imported into the R environment, "\<MRL" is converted to NA. Thus, we just need to use a conditional statement that converts NA's to 1/2 the MRL.

```{r, NAs, warning=FALSE, message=FALSE,error=FALSE}
#Read MLR from datasheet
mrl<-read_xlsx(
        path = paste0(data_dir,"nc_database.xlsx"), 
        sheet = 2, 
        n_max = 1, 
        col_types = c('text',rep('numeric',17)),
        col_names = df_names) %>%  
    pivot_longer(!Key, names_to="name", values_to = "mrl") %>% 
    select(name, mrl)

#Characterize the number of lead samples below 1ppb, 1-5 ppb, and >5ppb
mrl_numbers <- df %>% 
  pivot_longer(-Key) %>% 
  filter(name == 'Pb') %>% 
  mutate(
     n_1   = if_else(is.na(value), 1, 0), 
     n_1_5 = if_else(value>=1 & value<5, 1, 0),
     n_5   = if_else(value>=5, 1, 0)) %>% 
    group_by(name) %>% 
    summarise(
    n_1   = sum(n_1, na.rm = T),
    n_1_5 = sum(n_1_5, na.rm = T),
    n_5 = sum(n_5, na.rm = T)
  )

#Remove -9999 values from pH col
df <- df %>% mutate(pH = if_else(pH<0, NA_real_, pH))

#Now fill NAs with 1/2 the MRL value
df <- df %>% 
  pivot_longer(-Key) %>% 
  left_join(., mrl) %>% 
  mutate(value = if_else(is.na(value), mrl/2, value)) %>% 
  select(-mrl) %>% 
  pivot_wider(names_from = name, values_from = value)
```

## In-text calculations

```{r echo = FALSE, results = 'asis'}
 knitr::kable(mrl_numbers, caption = 'Number of lead samples below VT reporting limit (1 ppb), between VT and NC reporting limit (1-5ppb), and above the reporting imit (>5ppb)')
```

```{r echo = FALSE, results = 'asis'}
PbX<-df %>% 
  select(Pb, PbX) %>% 
  mutate(count = if_else(PbX>Pb, 1, 0)) %>% 
  summarise(sum = sum(count))
knitr::kable(PbX, caption = 'Number of samples where PbX > Pb')
```

# Step 3: Summary stats

```{r, summary stats, warning=FALSE, message=FALSE,error=FALSE}
#Summon the tidyverse!
summary_stats <- df %>% 
  #convert to long format
  pivot_longer(-Key) %>% 
  #Group by name
  group_by(name) %>% 
  #calculate summary stats
  summarise(
    mean   = mean(value, na.rm = T), 
    median = median(value, na.rm = T), 
    max    = max(value, na.rm = T), 
    min    = min(value, na.rm = T),
    q90    = quantile(value, 0.9, na.rm = T)
  ) 
```

```{r echo = FALSE, results = 'asis'}
 knitr::kable(summary_stats)
```

# Step 4: Fingerprinting

```{r warning=FALSE, message=FALSE,error=FALSE}
#Create function to run correlation
corr_fun<-function(n){
  
  #Define variables of interest
  var_1<-var$var_1[n]
  var_2<-var$var_2[n]
  
  #Select variables of interest
  temp <- df %>% select(var_1, var_2)
  #Run cor test
  temp <- cor.test(
            x=as_vector(temp[,1]),
            y=as_vector(temp[,2]), 
            method = 'spearman')
  
  #Extract information of interest and export
  tibble(
    var_1, 
    var_2, 
    p_value = temp$p.value, 
    rho = temp$estimate)
}

#Create tibble of variables to run correlation analysis on
var<-tibble(
  var_1 = c(rep("Pb", 5), rep('PbX', 6)), 
  var_2 = c('Zn', 'Cu', 'Cd', 'Fe', 'Sn', "ZnX", "CuX", "pH", "CdX", "FeX", "SnX")
)

#Apply function to variable list
rho_values <- 
  lapply(
    X = seq(1, nrow(var)), 
    FUN = corr_fun) %>% 
  bind_rows()
```

```{r echo = FALSE, results = 'asis'}
 knitr::kable(rho_values)
```

# Step 5: System Characteristics

## Well construction

```{r warning=FALSE, message=FALSE,error=FALSE}
#First, we need to download well construnction info
well_info <- 
  read_xlsx(
    path = paste0(data_dir,"nc_database.xlsx"), 
    sheet = 3)

#now summarise by construction
well_construction<-df %>% 
  left_join(., well_info) %>% 
  group_by(`System type`) %>% 
  summarise(
    n_wells    = n(),
    PbX_median = median(PbX), 
    pH_median  = median(pH, na.rm=T)
  )
```

```{r echo = FALSE, results = 'asis'}
 knitr::kable(well_construction)
```

## Well Depth

```{r warning=FALSE, message=FALSE,error=FALSE}
#Summarise by depth
well_depth <- df %>% 
  left_join(., well_info) %>% 
  group_by(`Catorgized well depth`) %>% 
  summarise(
    n_wells    = n(),
    PbX_median = median(PbX), 
    pH_median  = median(pH, na.rm=T)
  )
```

```{r echo = FALSE, results = 'asis'}
 knitr::kable(well_depth)
```

## Geology

```{r warning=FALSE, message=FALSE,error=FALSE}
#download location info
location_info <- 
  read_xlsx(
    path = paste0(data_dir,"nc_database.xlsx"), 
    sheet = 1)
  
#Summarise by geology
well_geology<- df %>% 
  left_join(., location_info) %>% 
  group_by(Geology) %>% 
  summarise(
    n_wells    = n(),
    PbX_median = median(PbX), 
    pH_median  = median(pH, na.rm=T)
  )
```

```{r echo = FALSE, results = 'asis'}
 knitr::kable(well_geology)
```

## Well construction and depth

```{r warning=FALSE, message=FALSE,error=FALSE}
#Summarise by well construction and depth
well_con_dep <- df %>% 
  left_join(., well_info) %>% 
  group_by(`System type`,`Catorgized well depth`) %>% 
  summarise(
    n_wells    = n(),
    PbX_median = median(PbX), 
    pH_median  = median(pH, na.rm=T)
  )
```

```{r echo = FALSE, results = 'asis'}
 knitr::kable(well_con_dep)
```

## Well construction and geology

```{r warning=FALSE, message=FALSE,error=FALSE}
#Summarise by well construction and geology
well_con_geo <- df %>% 
  left_join(., well_info) %>% 
  left_join(., location_info) %>% 
  group_by(Geology,`System type`) %>%   summarise(
    n_wells    = n(),
    PbX_median = median(PbX), 
    pH_median  = median(pH, na.rm=T)
  ) 
```

```{r echo = FALSE, results = 'asis'}
 knitr::kable(well_con_geo)
```

## Well depth and geology

```{r warning=FALSE, message=FALSE,error=FALSE}
#Summarise by well depth and geology
well_dep_geo <- df %>% 
  left_join(., well_info) %>% 
  left_join(., location_info) %>% 
  group_by(Geology,`Catorgized well depth`) %>%   
  summarise(
    n_wells    = n(),
    PbX_median = median(PbX), 
    pH_median  = median(pH, na.rm=T)
  ) 
```

```{r echo = FALSE, results = 'asis'}
 knitr::kable(well_dep_geo)
```

## Well construction, depth, and geology

```{r warning=FALSE, message=FALSE,error=FALSE}
#Summarise by well construction, depth, and geology
well_con_dep_geo <- df %>% 
  left_join(., well_info) %>% 
  left_join(., location_info) %>% 
  group_by(Geology,`System type`,`Catorgized well depth`) %>%   summarise(
    n_wells    = n(),
    PbX_median = median(PbX), 
    pH_median  = median(pH, na.rm=T)
  ) 
```

```{r echo = FALSE, results = 'asis'}
 knitr::kable(well_con_dep_geo)
```

# Step 6: Demographics

## Median Pb Levels

```{r warning=FALSE, message=FALSE,error=FALSE}
#Summarise by demographics
well_demographics <- df %>% 
  left_join(., well_info) %>% 
  group_by(Race) %>%
  summarise(
    n_wells    = n(),
    PbX_median = median(PbX), 
    pH_median  = median(pH, na.rm=T)
  ) 
```

```{r echo = FALSE, results = 'asis'}
 knitr::kable(well_demographics)
```

## Comparative stats

```{r echo=T, warning=F, message=F}
#create vectors of white and bipoc PbX values
BIPOC_PbX<-df %>% 
    left_join(., well_info) %>% 
    filter(Race =="BIPOC") %>% 
    select(PbX) %>% 
    as_vector()
white_PbX<-df %>% 
    left_join(., well_info) %>% 
    filter(Race =="White") %>% 
    select(PbX) %>% 
    as_vector()

#Run wilcox test
wilcox.test(BIPOC_PbX, white_PbX)
```
